---
title: "Global Poverty Trends (2016-2026)"
format: html
embed-resources: true
editor: visual
execute:
  warning: false
  message: false
---

Data Sources: Our World in Data - Poverty ([World Bank (2026). Poverty and Inequality Platform (version 20260324_2021 and 20260324_2017) \[Data set\]. World Bank Group. https://pip.worldbank.org/.](https://pip.worldbank.org/)) ; World Bank Global Economic Indicators (2010-2025)(<https://www.kaggle.com/datasets/tanishksharma9905/global-economic-indicators-20102025?resource=download>)

# Visualization 1

```{r}
library(tidyverse)
library(readr)
owi_poverty <- read_csv("data/share-of-population-living-in-extreme-poverty.csv") 

poverty_world <- owi_poverty |>
  filter(Country %in% c("World", "World (excluding China)", "World (excluding India)"))
  
ggplot(data = poverty_world,
       mapping = aes(x = Year,
                     y = `Share below $3 a day`,
                     color = Country)) +
  geom_point(size = 2, shape = 16) +
  geom_text(aes(label = round(`Share below $3 a day`, 1)),
            vjust = -0.7,
            size = 3) +
  geom_smooth(method = "loess",
              se = FALSE) +
  scale_x_continuous(breaks = seq(2016, 2026, by = 1)) +
  scale_y_continuous(breaks = seq(9.5, 15.5, by = 1)) +
  theme_minimal() +
  labs(title = paste("Share of global population below 3$ per day",
                     "\n3 different measures, 2016-2026"),
       subtitle = paste("China and India have proven to significantly ",
                        "influenced\nglobal poverty levels over the last",
                        "decade"),
       caption = "Source: World Bank Our World in Data Poverty Database")
  
  
  
```

This visualization tracks the share of people living \$3 a day for the world, the world excluding China, and the world excluding India over the past decade. The share of people living below \$3 a day has fallen over this time period, and the gap between the world and world excluding China trends shows the significant impact that China's poverty rate influences that of the world. The "excluding India" line shows how much the global average depends on India's trajectory in poverty reduction. Policy implications are, most notably, that looking at aggregate measures alone obscure uneven progress across countries. Policies and programs to end or significantly reduce poverty should thus target countries with the highest concentration of the world's remaining poor.

# Visualization 2

```{r}
library(tidyverse)
library(dplyr)

WB_econ_indicators <- read_csv("data/world_bank_data_2025.csv") 

#subset of countries: bottom 10 GDP per capita
bottom5_gdppc <- WB_econ_indicators |>
  group_by(year) |>
  slice_min(`GDP per Capita (Current USD)`, n = 5)

bottom5_gdppc_2023 <- bottom5_gdppc |>
  filter(year == 2023) |>
  mutate(
    region = case_when(
      country_name %in% c("Afghanistan", "Yemen, Rep.") ~ "Middle East",
      country_name == "Central African Republic" ~ "Central Africa",
      country_name %in% c("Burundi", "Madagascar") ~ "East Africa"
    ))

ggplot(data = bottom5_gdppc_2023,
       mapping = aes(x = country_name,
                     y = `GDP per Capita (Current USD)`,
                     fill = region)) +
  geom_col() +
  scale_x_discrete(guide = guide_axis(angle = 30)) +
  labs(x = "Country",
       title = "Bottom 5 Countries by GDP per Capita (in current USD), 2023",
       subtitle = paste("All 5 countries with the lowest GDP per capita",
                        "globally are located in Africa\nor the Middle East"),
       caption = "Source: World Bank Global Economic Indicators, 2010-2025") +
  theme_minimal()
  


```

This bar chart shows the five poorest countries by GDP per capita, which are concentrated in Africa and the Middle East (Madagascar, Burundi, the Central African Republic, Yemen, and Afghanistan). The main objective of this visualization is to show if and where poverty is concentrated globally, following the policy implications of the previous visualization. The beginning of the answer to this question, looking at this bar chart, is that poverty is heavily concentrated in Africa, primarily, and also the Middle East. If this visualization were to include the 20 poorest countries by GDP per capita, I imagine this would widen the scope of regions to look at, likely adding Southeast Asia and Latin America to the list. Implications for policy are most notably that, if we are to address global poverty, we should start by concentrating efforts on those regions at the bottom end of the income and consumption distribution.

# Visualization 3

```{r}
library(tidyverse)
library(ggrepel)

WB_econ_indicators_2023 <- WB_econ_indicators |>
  filter(year == 2023)

outliers <- WB_econ_indicators_2023 |>
  filter(
    `GDP Growth (% Annual)` < -5 |
    `GDP Growth (% Annual)` > 10 |
    `Unemployment Rate (%)` > 20
  )

ggplot(data = WB_econ_indicators_2023,
       mapping = aes(x = `GDP Growth (% Annual)`,
                     y = `Unemployment Rate (%)`,
                     size = `GDP per Capita (Current USD)`)) +
  geom_point(shape = 21,
             color = "white",
             fill = "red",
             alpha = 0.8) +
  geom_text_repel(data = outliers,
                  aes(label = country_name),
                  size = 2.5) +
  scale_size_area(name = "GDP per Capita (Current USD)",
                  max_size = 20) +
  labs(title = "Global Economic Growth, Unemployment, and Income, 2023",
       subtitle = paste("Outlier countries in either unemployment or GDP",
       "growth\nare labeled, point size represents GDP per capita"),
       caption = "Source: World Bank Global Economic Indicators, 2010-2025") +
  theme_minimal() +
  theme(
    legend.key.size = unit(0.5, "cm"),
    legend.text = element_text(size = 5),
    legend.title = element_text(size = 6)
  ) 

```

This bubble chart compares GDP growth, unemployment, and GDP per capita. Most countries cluster around low-to-moderate annual GDP growth and low unemployment. Since bubble size encodes GDP per capita, it is clear from this visualization that the wealthiest countries (biggest bubbles) sit in the stable middle, while small economies account for almost all of the outliers (all of which are labeled). The weak-to-nonexistent relationship between GDP growth and unemployment is the main policy implication of this visualization – economic growth alone does not necessarily translate into jobs – this last step requires accompanying and well thought-out labor-market policies.

# Visualization 4

```{r}
library(tidyverse)

WB_econ_indicators <- read_csv("data/world_bank_data_2025.csv")

bottom5 <- WB_econ_indicators |>
  filter(year == 2019) |>
  slice_min(`GDP per Capita (Current USD)`, n = 5) |>
  pull(country_name)

top5 <- WB_econ_indicators |>
  filter(year == 2019) |>
  slice_max(`GDP per Capita (Current USD)`, n = 5) |>
  pull(country_name)

covid_data <- WB_econ_indicators |>
  filter(
    year >= 2018,
    year <= 2023,
    country_name %in% c(bottom5, top5)
  ) |>
  mutate(
    group = case_when(
      country_name %in% bottom5 ~ "Bottom 5",
      country_name %in% top5 ~ "Top 5"
    )
  )

covid_summary <- covid_data |>
  group_by(year, group) |>
  summarize(
    avg_growth = mean(`GDP Growth (% Annual)`, na.rm = TRUE)
  )

ggplot(data = covid_summary,
       mapping = aes(x = year,
                     y = avg_growth,
                     color = group,
                     linetype = group)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3,
             alpha = 0.8) +
  geom_vline(xintercept = 2020,
             linetype = "dotted") +
  scale_linetype(guide = "none") +
  labs(
    x = "Year",
    y = "Average GDP Growth (% Annual)",
    color = "Country Group",
    title = paste("COVID-19's Economic Impact on the World's Wealthiest",
                  "and\nPoorest 5 Countries by GDP per Capita"),
    subtitle = paste("Policies which allowed the wealthiest 5 countries",
                     "to recover their pre-shock\naverage GDP growth",
                     "significantly more quickly than the poorest 5",
                     "was followed\nby a subsequent drop below",
                     "that of the poorest 5"),
    caption = "Source: World Bank Global Economic Indicators, 2010-2025") +
  theme_classic()
  
```

This line chart compares the average growth rates of the 5 wealthiest economies and the 5 poorest economies over the course of the Covid-19 pandemic period, with the beginning of the pandemic marked by the black dotted line. From the graph, we see that wealthy countries, with fiscal capacity to implement stimulus and develop/carry out vaccinations, rebounded much more quickly than poorer countries, which lacked the resources for either significant stimulus or widespread vaccine distribution without external/international help. The policy implication of this visualization is that poor countries lack the tools and resources to deal with shocks of this scale, which should be an argument in favor of international support from wealthy countries to poorer ones in times of global shocks, which proved to be effective, if imperfect, for the Covid-19 shock.
